--- 
title: "Notes on 'Deep Learning with R'"
author: "Indrajeet Patil"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "These are my notes while reading the book `Deep Learning with R` by Francois Chollet and J.J.Allaire"
---

# Prerequisites

Before starting with the examples included in this book, I found it useful to
first clear up few basic concepts needed for data representation in R.

## Tensors

Deep learning involves data processing through layers of neural networks. So, of
course, the first question is what's the data structure and is it supported in
`R`? The basic data structure in machine learning world is **tensor**, which is
a generalization of vectors and matrices to higher dimensions (**D**). Tensors can be
constructed in `R` using `array` function.

There are no *scalars* (or **0D** tensors) in `R`, but a vector of length one is
conceptually similar to a scalar.

```{r scalar}
# creating the data structure
x0 <- 2

# checking its attributes
str(x0)
dim(as.array(x0))
typeof(x0)
```

Vectors have a single dimension (or they are **1D** tensors).

```{r vector}
# creating the data structure
x1 <- c(2, -3, 5, 66)

# checking its attributes
str(x1)
dim(as.array(x1))
typeof(x1)
```

Note that both scalars and vectors are one-dimensional arrays in `R`.

Matrices have two dimensions (or they are **2D** tensors).

```{r matrix}
# creating the data structure
x2 <- matrix(c(1:10), nrow = 2, ncol = 5)

# checking its attributes
str(x2)
dim(as.array(x2))
typeof(x2)
```

If we were to imagine a cube with a collection of numbers along each of its axes, that's what a **3D** tensor would like.

```{r tensor3d}
# creating the data structure
x3 <- array(data = c(1:24), dim = c(2, 3, 4))

# what it looks like
# (it's like stacking 2 x 3 matrices on top of each other 4 times!)
x3

# checking its attributes
str(x3)
dim(x3)
typeof(x3)
```

Each tensor has three key attributes-

  - *rank* (or number of dimensions/axes): `x3` is a **3D** tensor so it has a
    rank of `r length(dim(x3))`.
  - *shape*: Confusingly, this refers to the number of dimensions (`dim()`)
    along each of these axes. For `x3`, this was `r dim(x3)`.
  - *type*: Type of data contained in the tensor. For `x3`, this was
    `r typeof(x3)`.

So far we have been looking at made-up examples of tensors. Let's have a look at
the kind of data we will encounter in a typical ML workflow. We will look at the
MNIST dataset contained in the `keras` library.

```{r mnist}
library(keras)
use_condaenv("r-tensorflow")

mnist <- keras::dataset_mnist()
str(mnist)
```
